---
title: "Machine Learning - Course Project"
author: "Daniel Ramírez Torres"
date: "22/08/2015"
output: html_document
---

## Introduction

In this project, we will use the information obtained from several accelerometers of 6 participants. The participants were asked to perform barebell lifts correctly an dincorrectly in 5 different ways. The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set.

In the end, the question to answer is: can we predict the quality of the activity?


## Loading and exploring the data

First we need to download our training and test sets, as well as load the caret package which will be used later.

Training set download: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

Test set download: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


```{r}
library(caret)
dtrain <- read.csv("pml-training.csv")
dtest <- read.csv("pml-testing.csv")
```

As we want to build a predictor, we'll decompose our training set into two datasets (one for testing and the other to use as supervised training)

```{r}
set.seed(1234)
split <- createDataPartition(y=dtrain$classe, p=0.8, list=FALSE)
training <- dtrain[split, ]
testing <- dtrain[-split, ]
ncol(training)
```


As we can see, we have a lot of features in our dataset. In the next section we will analyze them and clean the dataset in order to use only the relevant features.

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.


## Cleaning the dataset

As we want to reduce the number of features (in both training and testing), we'll do it by removing variables with:

- No intuitive sense in this prediciton.
- Zero variance (or almost zero).
- Most common value NA.

```{r}
# Removing the first 5 columns: X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp) as they dont make sense for prediction.
training <- training[, -(1:5)]
testing <- testing[, -(1:5)]

# Removing zero variance values
indexZero <- nearZeroVar(testing)
training <- training[, -indexZero]
testing <- testing[, -indexZero]

# Removing variables with NA as most common value.
indexNA <- sapply(training, function(x) mean(is.na(x))) > 0.9
training <- training[, indexNA==FALSE]
testing <- testing[, indexNA==FALSE]
ncol(training)
```

## Modeling with Random Forests

As I have some experience with random forests, I have decided to create my model using them. I expect a high accuracy althoug not really good performance (as random forest is quite expensive). Let's build the model:

```{r}
library(randomForest)
model <- randomForest(classe ~. , data=training)
model
```

Now we can predict our in-sample error using the predict function:

```{r}
predictions <- predict(model, testing, type="class")
```

Finally, we can test our results using a confusion matrix:

```{r}
confusionMatrix(predictions, testing$classe)
```

As our accuracy is 99.8%, our predicted accuracy for the out-of-sample error is 0.2%.

## Generating testfiles

```{r}
output <- predict(model, dtest, type="class")
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(output)
```